Categorical EGraphs for Lambda Calculi

### 2025-07-11 POPL2026/notes/SlottedNotes.tex
fixed figures, updated contributions
some fixes

Lambda calculus with explicit substitution (also $\lambda_{\sigma}$-calculus) made the substitution operator a first-class element of the calculus (rather than a meta-level operation) to bridge the gap between lambda calculus and its concrete implementations~\cite{Abadi_Cardelli_Curien_Lévy_1991}.
However, the existence of unique normal forms turned out to be a delicate matter and to achieve uniqueness, lambda terms must be quotiented by \textit{graphical} equivalence in addition to $\alpha$-equivalence~\cite{Standardisation}.
We claim that using e-hypergraphs over e-graphs (or slotted e-graphs) may be more beneficial for doing equational reasoning for lambda calculus with explicit substitution.

@@

\begin{tabular}{lcc}
\texttt{subst\_same\_var} & $x[x / u] = u$ &\\
\texttt{subst\_diff\_var} & $y[x / u] = y$ & if $x \not \in \mathcal{F}(u)$\\
\texttt{subst\_application} & $(uu')[x / v] = u[x / v]u'[x / v]$ & \\
\texttt{subst\_lambda} & $(\lambda x . u)[y / v] = \lambda x . (u[y / v])$ & if $x \not \in \mathcal{F}(v)$ \\
\texttt{subst\_interchange\_1} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v[x / v']]$ & if $x' \in \mathcal{F}(v)$ and $x \in \mathcal{F}(v')$\\
\texttt{subst\_interchange\_2} & $(u[x / v])[x' / v'] = (u[x / v[x' / v']])[x / v[x / v']]$ & if ???\\
\texttt{subst\_comm} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v]$ & if $x \not \in \mathcal{F}(v')$ and $x' \not \in \mathcal{F}(v)$\\
\end{tabular}

>>>>

Lambda calculus with explicit substitution (also $\lambda_{\sigma}$-calculus) made the substitution operator a first-class element of the calculus (rather than a meta-level operation) to bridge the gap between lambda calculus and its concrete implementations~\cite{AbadiExplicitSubst}.
However, the existence of unique normal forms turned out to be a delicate matter and to achieve uniqueness, lambda terms must be quotiented by \textit{graphical} equivalence in addition to $\alpha$-equivalence~\cite{Standardisation}.
We claim that using e-hypergraphs over e-graphs (or slotted e-graphs) may be more beneficial for doing equational reasoning for lambda calculus with explicit substitution.

@@

\begin{tabular}{lcc}
\texttt{beta} & $(\lambda x . t) u = t[x / u]$&\\
\texttt{subst\_same\_var} & $x[x / u] = u$ &\\
\texttt{subst\_diff\_var} & $y[x / u] = y$ & if $x \not \in \mathcal{F}(u)$\\
\texttt{subst\_application} & $(uu')[x / v] = u[x / v]u'[x / v]$ & \\
\texttt{subst\_lambda} & $(\lambda x . u)[y / v] = \lambda x . (u[y / v])$ & if $x \not \in \mathcal{F}(v)$ \\
\texttt{subst\_interchange\_1} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v[x' / v']]$ & if $x' \in \mathcal{F}(v)$ and $x \in \mathcal{F}(v')$\\
\texttt{subst\_interchange\_2} & $(u[x / v])[x' / v'] = (u[x / v[x' / v']])$ & if $x' \not \in \mathcal{F}(u)$\\
\texttt{subst\_comm} & $(u[x / v])[x' / v'] = (u[x' / v'])[x / v]$ & if $x \not \in \mathcal{F}(v')$ and $x' \not \in \mathcal{F}(v)$\\
\end{tabular}

### 2025-07-11 POPL2026/paper/sections/introduction.tex
some fixes

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest, particularly for optimisation and synthesis using the technique of \emph{equality saturation}~\cite{10.1145/1594834.1480915, flatt_small_2022, EggPaper,flatt_small_2022}, wherein rewrites to an e-graph are repeated until a fixpoint (or timeout) is reached.
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
We are interested in using e-graphs in the context of functional programming languages, which builds upon an internal representation of the $\lambda$-calculus.
Pioneering work in this direction has been done by~\citet{koehler2022sketchguided}, where the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.

@@

A simple extension of $\lambda$-calculus is given by adding \emph{explicit substitution}, which models \textbf{let}-bindings in programming languages.
Here, a term equipped with explicit substitutions $t[y/v][x/u]$ (modelling $\textbf{let} \; x = u \; \textbf{in} \; \textbf{let} \; y = v \; \textbf{in} \; t$) can be identified with the corresponding term where the order of substitution is reversed $t[x/u][y/v]$, supposing the substitutions are non-interacting.
This imposes a \emph{graphical equivalence}~\cite{accattoli2014nonstandard} of terms.
Modelling this in the conventional language of e-graphs requires the introduction of additional nodes to represent the explicit substitutions, and additional rewrite rules which capture graphical equivalence.
This pollutes the state space of equality saturation for these languages: the interesting rewriting is that of the user-specified algebraic theory one wishes to model, not between semantically equivalent but syntactically distinct forms in the language.

All these issues can be resolved by the use of \emph{string diagrams} in place of textual syntax.
\citet{tiurin2025equivalencehypergraphsdporewriting}~shows how e-graphs (without binding) can be given a categorical semantics using symmetric monoidal semilattice-enriched categories, which allows e-graphs to be reconstructed as string diagrams.
In this paper, we extend this technique to e-graphs with bindings, lifting to monoidal \emph{closed} categories.

@@

String diagrams are a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, which are a fundamental data structure in algorithm design.
These categorical techniques provide insight in how to solve the problems of binding and capture-avoiding substitution in e-graphs, as we develop in this paper.

The general background literature for string diagrams and their relationship to formal languages is summarised by~\citet{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams with bindings is given in~\citet{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\citet{tiurin2025equivalencehypergraphsdporewriting}.

String diagrams provide a topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out.
In this article, our string diagrams are oriented left-to-right, bottom-to-top to make the pictures look more similar to those commonly drawn for e-graphs.
The structure of symmetric monoidal categories is absorbed into the graphical syntax, with the composition and tensoring represented by juxtaposition, and symmetries as weaving of strings.
This visual approach hides the bureacracy in complex operations and reveals insights sometimes obscured by traditional one-dimensional symbolic notation.
Originating from notations by~\citet{hotzsd} (for automata) and~\citet{penrose1984spinors} (for tensor networks), extensions of string diagrammatic calculi exist for various forms of monoidal categories and even bicategories~\cite{Selinger_2010}, and provide a mathematically rigorous reasoning toolkit~\cite{joyal_geometry_1991}.

Commonly, string diagrams are used informally to provide a two-dimensional syntax for the internal language of a category, but for our purposes we must be more formal.
Specifically, we require a concrete encoding of string diagrams, analogous to the de Bruijn index representation of $\lambda$-terms.
For this, we use \emph{hypergraphs}~\cite{bonchi_string_2022-1}, which have nice syntactic characteristics.
For instance, hypergraph isomorphism captures when string diagrams \enquote{look the same} --- topologically, that one may be deformed continuously into the other without obstruction\footnote{This is the informal notion of the \enquote{isotopy of string diagrams}.}; secondly, string-diagrammatic equational reasoning can be formalised as (categorical) graph rewriting over these hypergraphs using the technique of double-pushout rewriting (DPO)~\cite{bonchi_string_2022-1}.

Finally, for the diagrammatic representation of two concepts essential to both programming languages (variable binding) and to e-graphs (equivalence classes) we rely on the concept of \emph{functorial boxes} as introduced by Melli\`es~\cite{10.1007/11874683_1}.
The diagrammatic boxes require the introduction of further structure to hypergraphs in the form of a hierarchical layering of the graphs (\emph{hierarchical hypergraphs})~\cite{fscd}.

@@

Revisiting the earlier example $\lambda f . f ((\lambda x . f x) 2)$, in~\autoref{fig:de-brujin-string}, observe that the two occurrences of $f$ are now shared.
$\lambda$-abstraction is represented by a rounded box with the bound variable a dangling wire exiting the frame along its vertical boundary, the application node a half-circle
\adjustbox{scale=0.15}{\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=none] (0) at (1, 6.75) {};
			\node [style=none] (1) at (1.25, 7.75) {};
			\node [style=none] (2) at (2, 8) {};
			\node [style=none] (3) at (2.75, 7.75) {};
			\node [style=none] (4) at (3, 6.75) {};
			\node [style=none] (5) at (1.5, 6.75) {};
			\node [style=none] (6) at (2.5, 6.75) {};
			\node [style=none] (7) at (2, 9) {};
			\node [style=none] (8) at (1.5, 5.75) {};
			\node [style=none] (9) at (2.5, 5.75) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw [style=lambda_unit_string] (3.center)
			to [in=60, out=-30, looseness=0.75] (4.center)
			to (0.center)
			to [in=-150, out=120, looseness=0.75] (1.center)
			to [in=-180, out=45, looseness=0.75] (2.center)
			to [in=135, out=0, looseness=0.75] cycle;
			\draw (2.center) to (7.center);
			\draw (5.center) to (8.center);
			\draw (9.center) to (6.center);
		\end{pgfonlayer}
	\end{tikzpicture}
},
and sharing a \emph{copy node} \comonoid.

@@

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the \emph{e-hypergraphs} of~\citet{tiurin2025equivalencehypergraphsdporewriting}.
For the initial term, $(a*2)/2$, which is already represented efficiently in the e-graph by sharing the node $2$, the string diagram version shares this node explicitly, using a copy node.

E-graphs consist of two types of nodes: \emph{e-classes}, representing equivalence classes of terms and denoted by dashed boxes, containing \emph{e-nodes}, representing term formers.
E-nodes reference e-classes instead of other e-nodes, in contrast to a traditional term graph.
By making a sequence of choices for e-nodes in e-classes, we can extract terms from the graph; intuitively, a different sequence of choices results in an equivalent term.
A term graph is embedded as an e-graph by an e-graph where each e-class has a single e-node.
The first rewrite rule, replacing multiplication by \emph{shift-left} creates the first non-singleton e-class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) nodes, now sharing the node $a$.
This second e-graph illustrates the other representational difference, besides sharing, between conventional e-graphs and e-hypergraphs.
In the former, edges connect directly to e-nodes inside the e-class, whereas in the latter edges connect to the equivalence class itself.
Explicit deletion nodes are depicted with black dots with a single incoming wire
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
inside the dashed boxes, indicating which of the equivalence-class-level edges are connected to which nodes inside the equivalence class.

@@

In slotted e-graphs, every e-class is parameterised by a set of free variables of the terms it represents.
The edges then additionally convey the information about how variables are instantiated.
For example, in~\autoref{fig:slotted}, an e-class containing $+$ that is parameterised by variable names $s_2$ and $s_3$ gets instantiated once with $a,b$ (via $s_4 \mapsto a, s_2 \mapsto s_4$ and $s_5 \mapsto b, s_3 \mapsto s_5$) and once with $x,c$.
The fact that the variable $x$ is bound is evident from the corresponding e-class being parameterised by only three variables.
If we were to add another e-node such as $e + d$, there would be an edge added to an already existing e-class such that $s_1 \mapsto e$, $s_2 \mapsto d$.
We can achieve a similar degree of sharing (in the sense that we have the same number of nodes for $+$) using string diagrams by performing a closure conversion of $+$ (\autoref{fig:closed}).
Fundamentally, there is a difference in approach here: slotted e-graphs solve the problem of variable binding generically using nominal techniques, in a way that can be applied to arbitrary formal languages with binders (e.g. $\pi$-calculus), while ours utilises an embedding into the internal language of our categorical model, which resembles a kind of metatheoretic $\lambda$-calculus, for which we use the metatheoretic $\lambda$-abstraction to handle the binding of variables.
We defer a case study to~\autoref{sec:application}, where we argue that our approach has advantages if one is particularly interested in e-graphs for (variations on) $\lambda$-calculus, and, by extension, programming languages.
However, these constructions are not unrelated: our categorical model can help justify the congruences maintained by slotted e-graphs, which we further elaborate on in~\autoref{sec:categorical}.
We also note that there is also a difference in the scopes of these related works: \citet{slotted-egraphs} focus on the e-graph as a data structure, providing a full implementation of their slotted e-graphs, while we focus on the foundational categorical semantics of e-graphs with binding and their combinatorial representation.

>>>>

E-graphs were first introduced in the automated theorem proving literature in the 1980s~\cite{nelson1980techniques}, but recent years have seen renewed interest, particularly for optimisation and synthesis using the technique of \emph{equality saturation}~\cite{10.1145/1594834.1480915, flatt_small_2022, EggPaper,flatt_small_2022}, wherein rewrites to an e-graph are repeated until a fixpoint (or timeout) is reached.
The wide range of application domains led to extensions of the e-graph formalism, for instance to account for conditional rewriting~\cite{singher2023colored} or  to combine e-graphs with database queries (\emph{relational e-matching})~\cite{zhang_relational_2022}.
We are interested in using e-graphs in the context of functional programming languages, which builds upon an internal representation of the $\lambda$-calculus.
% Pioneering work in this direction has been done by~\citet{koehler2022sketchguided}, where the authors outlined the two interconnected challenges of binding and capture-avoiding substitution that arise when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.
\citet{koehler2022sketchguided} pioneered work in this area by identifying two key, interconnected challenges—handling binding and ensuring capture-avoiding substitution—when applying equality saturation to an encoding of the $\lambda$-calculus in e-graphs.

@@

% A simple extension of $\lambda$-calculus is given by adding \emph{explicit substitution}, which models \textbf{let}-bindings in programming languages.
A simple but theoretically~\cite{AbadiExplicitSubst} and practically~\cite{anf} important extension of the $\lambda$-calculus consists of adding let-bindings or explicit substitutions to the syntax.
% Here, a term equipped with explicit substitutions $t[y/v][x/u]$ (modelling $\textbf{let} \; x = u \; \textbf{in} \; \textbf{let} \; y = v \; \textbf{in} \; t$) can be identified with the corresponding term where the order of substitution is reversed $t[x/u][y/v]$, supposing the substitutions are non-interacting.
Terms are usually identified up to permutations of non-interacting substitutions, which is sometimes called \enquote{graph equivalence}~\cite{accattoli2014nonstandard} since such identification becomes obvious in certain diagrammatic representations.
% This imposes a \emph{graphical equivalence}~\cite{accattoli2014nonstandard} of terms.
Modelling this in the conventional language of e-graphs requires the introduction of additional nodes to represent the explicit substitutions, and additional rewrite rules which capture graphical equivalence.
This pollutes the state space of equality saturation for these languages: the interesting rewriting is that of the user-specified algebraic theory one wishes to model, 
% not between semantically equivalent but syntactically distinct forms in the language.
not the bureaucracy of managing the congruences of semantically equivalent yet syntactically distinct forms in the language.
All these issues can be resolved by the use of \emph{string diagrams} in place of textual syntax.
\citet{tiurin2025equivalencehypergraphsdporewriting}~shows how e-graphs (without binding) can be given a categorical semantics using symmetric monoidal semilattice-enriched categories, which allows e-graphs to be reconstructed as string diagrams.
In this paper, we extend this technique to e-graphs with bindings, lifting to monoidal \emph{closed} categories.

@@

String diagrams are a direct bridge from the language of category theory, which in turn can be used to give semantic models to a variety of programming languages and logics, to the concrete language of graphs, which are a fundamental data structure in algorithm design.
These categorical techniques provide methodological guidance in how to solve the problems of binding and capture-avoiding substitution in e-graphs, as we develop in this paper.

The general background literature for string diagrams and their relationship to formal languages is summarised by~\citet{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams for $\lambda$-calculi is given in~\citet{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\citet{tiurin2025equivalencehypergraphsdporewriting}.

String diagrams provide a topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out.
In this article, our string diagrams are oriented left-to-right, bottom-to-top to make the pictures look more similar to those commonly drawn for e-graphs.
The structure of symmetric monoidal categories is absorbed into the graphical syntax, with the composition and tensoring represented by juxtaposition, and symmetries as weaving of strings.
This visual approach hides the bureacracy of complex operations and reveals insights sometimes obscured by traditional one-dimensional (textual) syntax.
Originating from notations by~\citet{hotzsd} (for automata) and~\citet{penrose1984spinors} (for tensor networks), extensions of string diagrammatic calculi exist for various forms of monoidal categories and even bicategories~\cite{Selinger_2010}, and provide a mathematically rigorous reasoning toolkit~\cite{joyal_geometry_1991}.

Commonly, string diagrams are used informally to provide a two-dimensional syntax for the internal language of a category, but for our purposes we must be more formal.
Specifically, we require a concrete encoding of string diagrams, analogous to the de Bruijn index representation of $\lambda$-terms.
For this, we use \emph{hypergraphs}~\cite{bonchi_string_2022-1}, which have suitable theoretical properties.
For instance, hypergraph isomorphism rigorously expresses what it means for string diagrams to \enquote{look the same} --- topologically, that one may be deformed continuously into the other without obstruction\footnote{This is the informal notion of the \enquote{isomorphism of string diagrams}.}; secondly, string-diagrammatic equational reasoning can be formalised as (categorical) graph rewriting over these hypergraphs using the technique of double-pushout rewriting (DPO)~\cite{bonchi_string_2022-1}.
Finally, for the diagrammatic representation of two concepts essential to both programming languages (variable binding) and to e-graphs (equivalence classes) we rely on the concept of \emph{functorial boxes} as introduced by Melli\`es~\cite{10.1007/11874683_1}.
The diagrammatic boxes require the introduction of further structure to hypergraphs in the form of a hierarchical layering of the graphs (\emph{hierarchical hypergraphs})~\cite{fscd}.

@@

Revisiting the earlier example $\lambda f . f ((\lambda x . f x) 2)$, in~\autoref{fig:de-brujin-string}, observe that the two occurrences of $f$ are now shared.
The scope of a $\lambda$-abstraction is represented by a rounded box with the bound variable represented as a wire attached to the frame, the application node a half-circle
\adjustbox{scale=0.15}{\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=none] (0) at (1, 6.75) {};
			\node [style=none] (1) at (1.25, 7.75) {};
			\node [style=none] (2) at (2, 8) {};
			\node [style=none] (3) at (2.75, 7.75) {};
			\node [style=none] (4) at (3, 6.75) {};
			\node [style=none] (5) at (1.5, 6.75) {};
			\node [style=none] (6) at (2.5, 6.75) {};
			\node [style=none] (7) at (2, 9) {};
			\node [style=none] (8) at (1.5, 5.75) {};
			\node [style=none] (9) at (2.5, 5.75) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw [style=lambda_unit_string] (3.center)
			to [in=60, out=-30, looseness=0.75] (4.center)
			to (0.center)
			to [in=-150, out=120, looseness=0.75] (1.center)
			to [in=-180, out=45, looseness=0.75] (2.center)
			to [in=135, out=0, looseness=0.75] cycle;
			\draw (2.center) to (7.center);
			\draw (5.center) to (8.center);
			\draw (9.center) to (6.center);
		\end{pgfonlayer}
	\end{tikzpicture}
},
and sharing a \emph{copy node} \comonoid.

@@

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the \emph{e-hypergraphs} used by~\citet{tiurin2025equivalencehypergraphsdporewriting}.
For the initial term, $(a*2)/2$, which is already represented efficiently in the e-graph by sharing the node $2$, the string diagram version shares this node explicitly, using a copy node.

E-graphs consist of two types of nodes: \emph{e-classes}, representing equivalence classes of terms and denoted by dashed boxes, containing \emph{e-nodes}, representing term formers.
E-nodes reference e-classes instead of other e-nodes, in contrast to a traditional term graph.
By making a sequence of choices for e-nodes in e-classes, we can extract terms from the graph; intuitively, a different sequence of choices results in an equivalent term.
A term graph is embedded as an e-graph by an e-graph where each e-class has a single e-node.
The first rewrite rule, replacing multiplication by \emph{shift-left} creates the first non-singleton e-class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) nodes, now sharing the node $a$.
This second e-graph illustrates the other representational difference\footnote{Despite the \enquote{difference}, these representations are isomorphic is a sense that there is a non-ambiguous translation from one representation to another and vice versa.}, besides sharing, between conventional e-graphs and e-hypergraphs.
In the former, edges connect directly to e-nodes inside the e-class, whereas in the latter edges connect to the equivalence class itself.
Explicit deletion nodes are depicted with black dots with a single incoming wire
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
inside the dashed boxes, indicating which of the equivalence-class-level edges are connected to which nodes inside the equivalence class.

@@

In slotted e-graphs, every e-class is parameterised by a set of free variables of the terms it represents.
The edges then additionally convey the information about how variables are instantiated.
For example, in~\autoref{fig:slotted}, an e-class containing $+$ that is parameterised by variable names $s_2$ and $s_3$ gets instantiated once with $a,b$ (via $s_4 \mapsto a, s_2 \mapsto s_4$ and $s_5 \mapsto b, s_3 \mapsto s_5$) and once with $x,c$.
The fact that the variable $x$ is bound is evident from the corresponding e-class being parameterised by only three variables.
If we were to add another e-node such as $e + d$, there would be an edge added to an already existing e-class such that $s_1 \mapsto e$, $s_2 \mapsto d$.
We can achieve a similar degree of sharing (in the sense that we have the same number of nodes for $+$) using string diagrams by performing a closure conversion of $+$ (\autoref{fig:closed}).
% One can consider the example from~\autoref{fig:e-graph-substitution} as an instance of a slotted e-graph as well (with headers omitted), the explicit substitution works similarly between the two versions of e-graphs with the slotted version providing sharing up-to renaming.
A slotted e-graph version of~\autoref{fig:e-graph-substitution} would be able to introduce sharing for the variable names $x$ and $y$ but must still represent abstraction, application, and substitution explicitly, and so would look much the same.
Fundamentally, there is a difference in approach here: slotted e-graphs solve the problem of variable binding generically using nominal techniques, in a way that can be applied to arbitrary formal languages with binders (e.g. $\pi$-calculus), while ours utilises an embedding into the internal language of our categorical model, which resembles a kind of metatheoretic $\lambda$-calculus, for which we use the metatheoretic $\lambda$-abstraction to handle the binding of variables.
We defer a case study to~\autoref{sec:application}, where we argue that our approach has advantages if one is particularly interested in e-graphs for (variations on) $\lambda$-calculus, and, by extension, programming languages.
However, these constructions are not unrelated: our categorical model can help justify the congruences maintained by slotted e-graphs, which we further elaborate on in~\autoref{sec:categorical}.
We also note that there is also a difference in the scopes of these related works: \citet{slotted-egraphs} focus on the e-graph as a data structure, providing a full implementation of their slotted e-graphs, while we focus on the foundational categorical semantics of e-graphs with binding and their combinatorial representation.
